
# Chapter 4 - Workflow: basics {-}

```{r message=F, error=F, warning=F}
library(tidyverse)
```

## **4.4** Practice {-}

### 1. Why does this code not work? {-}

```{r eval=FALSE}
my_variable <- 10
my_varıable
```

The code does not work because there is a typo in the variable name that you are calling. The letter "i" is not the same in my_var(i)able.

### 2. Tweak each of the following R commands so that they run correctly: {-}

```{r eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

fliter(mpg, cyl = 8)
filter(diamond, carat > 3)
```

To fix the commands, change "dota" to "data", "fliter" to "filter", "=" to "==", and "diamond" to "diamonds".

```{r chap4_exercises}
library(tidyverse)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

filter(mpg, cyl == 8)
filter(diamonds, carat > 3)
```

### 3. Press Alt + Shift + K. What happens? How can you get to the same place using the menus? {-}

This opens up a list of the keyboard shortcuts! Very useful. Using the menus, either type "shorcut" into the search bar under help, or find it under: tools: keyboard shortcuts help.

