---
title: "R for Data Science Walkthrough Chapters 9-13"
author: "Erick Lu"
output: 
  github_document:
    html_preview: false
    toc: true
    toc_depth: 2
    fig_width: 5
    fig_height: 3.5
---

This my walkthrough for chapters 9-13 for the book: _R for Data Science_ by Hadley Wickham and Garrett Grolemund. Here I provide solutions to their exercises and some of my own notes and explorations.

# Chapter 9 Introduction

This chapter provides a list of the content covered in chapter 10-16, which includes tibbles, data import, tidy data, relational data, strings (regular expressions), factors, and dates/times.

```{r}
library(tidyverse)
```


# Chapter 10 Tibbles

You can create tibbles from existing data frames using as_tibble(), or create brand new tibbles using tibble():

```{r}
tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
)
```

A cousin of tibble(), tribble(), can also be used as a way to manually enter data into a tibble format:

```{r}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```

You can also use non-syntactic names for variables in tibbles:

```{r}
tb <- tibble(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number"
)
tb
```

When compared to a data.frame in baseR, the tibble looks more user-friendly. Calling a tibble automatically provides only the beginning chunk of the data rather than filling up your entire console (think if it as default head(data.frame) display). Other nice features include not converting strings to factors or changing variable names.

To convert tables to or from data frames, use as_tibble() and as.data.frame():

```{r}
class(iris)
class(as_tibble(iris))
class(as.data.frame(as_tibble(iris)))
```

You can select columns in tibbles the same way you would with a data.frame:

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)
# extract column 'x' using either $ or [[]]
df$x
df[["x"]]
df[[1]]
```

## 10.5 Exercises

### 1. How can you tell if an object is a tibble? (Hint: try printing mtcars, which is a regular data frame).

You can tell if an object is a tibble because the output you get by calling it will say "tibble"! For example, calling the diamonds tibble returns :```# A tibble: 53,940 x 10``` as the first line of the output. Also you can tell something is a tibble based on the class specifications underneath each variable name. A tibble will also only print out the first 10 rows by default, whereas a data.frame will print out as many as the console allows. Last, the definitive way to tell something is a tibble is to use the class() function.

```{r}
class(diamonds)
class(mtcars)
```

### 2. Compare and contrast the following operations on a data.frame and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?

On a data.frame, ```df$x``` will still return the values for column xyz. This behavior does not occur for a tibble, which requires the exact name of the column ```df$xyz```. This data.frame feature might cause frustration if you have columns in your dataset with the same prefix, in which you might fetch the wrong column. The other functions between data.frame and tibble work the same way. One distinction to note is that, when creating the data.frame, "a" is considered a factor with 1 level. When creating the tibble, "a" is not converted into a factor.

```{r}
df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]

df <- tibble(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```

### 3. If you have the name of a variable stored in an object, e.g. var <- "mpg", how can you extract the reference variable from a tibble?

If the name of the variable is stored in an object, you can pass the object in lieu of the variable name using [[]] or [] just as you would do so with the explicit variable name. You can even pass the object and another variable name to obtain multiple reference variables using c(). I provide an example below using the diamonds dataset.

```{r}
var <- "carat"
var2 <- c("carat","price")

# extract only carat
diamonds[,var]

#extract carat and price
diamonds[,c(var,"price")]
diamonds[,var2]
```

### 4. Practice referring to non-syntactic names in the following data frame by:
```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
```

* Extracting the variable called 1.
```{r}
annoying[,"1"]
```
* Plotting a scatterplot of 1 vs 2.
```{r}
ggplot(annoying, aes(`1`,`2`))+
  geom_point()
```

* Creating a new column called 3 which is 2 divided by 1.
```{r}
annoying %>%
  mutate(`3` = `2`/`1`)
```

* Renaming the columns to one, two and three.
```{r}
annoying %>%
  mutate(`3` = `2`/`1`) %>%
  rename(one = `1`, two = `2`, three = `3` )
```


### 5. What does tibble::enframe() do? When might you use it?

Taken from the documentation: "enframe() converts named atomic vectors or lists to two-column data frames. For unnamed vectors, the natural sequence is used as name column." I might use this when I have a vector that I want to turn into a data.frame for graphing using ggplot, which requires data be in data.frame or tibble.

```{r}
x = rnorm(100)
names(x) <- c(5:104)
enframe(x)
class(enframe(x))
```

### 6. What option controls how many additional column names are printed at the footer of a tibble?

The documentation for ?format.tbl (tibble formatting) says that the n_extra argument will control how many additional columns to print abbreviated information for. The example provided in the documentation is below, which only prints 2 of the additional columns (whereas the unmodified print(flights) would yield 5 additional columns in the footer).

```{r}
print(nycflights13::flights, n_extra = 2)
```


# Chapter 11 Data Import

## 11.2 Getting started - reading in files

To practice various utilities for reading in files, we can use inline csv designation, which requires proper newline designation. Below are some examples of reading in inline csv chunks with various arguments tailored for the type of data being read in.

```{r}
# basic read_csv()
read_csv("a,b,c
         1,2,3 
         4,5,6")
# read in data, ignoring metadata lines
read_csv("The first line of metadata
The second line of metadata
  x,y,z
  1,2,3", skip = 2)
# designate lines to skip that start with specific symbol
read_csv("# A comment I want to skip
  x,y,z
  1,2,3", comment = "#")
# read in file that doesnt have column names
read_csv("1,2,3\n4,5,6", col_names = FALSE)
# read in file and specify column names
read_csv("1,2,3\n4,5,6", col_names = c("x", "y", "z"))
# read in file, replacing symbol with missing values (NA)
read_csv("a,b,c\n1,2,.", na = ".")
```

## 11.2.2 Exercises

### 1. What function would you use to read a file where fields were separated with “|”?

I would use ```read_delim()``` to read in a file where fields are separated with "|". For example:

```{r}
read_delim ("a|b|c
            1|2|3
            4|5|6", "|")
```

### 2. Apart from file, skip, and comment, what other arguments do read_csv() and read_tsv() have in common?

Based on the documentation, read_csv() and read_tsv() have col_names, col_types, locale, na, quoted_na, quote, trim_ws, n_max, guess_max, and progress.

### 3. What are the most important arguments to read_fwf()?

The most important arguments are the file, and the col_positions arguments. There are many options to specify col_positions, including fwf_empty(), fwf_widths(), fwf_positions(), and fwf_cols(). Below is the example provided in the documentation:

```{r}
fwf_sample <- readr_example("fwf-sample.txt")
cat(read_lines(fwf_sample))

# You can specify column positions in several ways:
# 1. Guess based on position of empty columns
read_fwf(fwf_sample, fwf_empty(fwf_sample, col_names = c("first", "last", "state", "ssn")))
# 2. A vector of field widths
read_fwf(fwf_sample, fwf_widths(c(20, 10, 12), c("name", "state", "ssn")))
# 3. Paired vectors of start and end positions
read_fwf(fwf_sample, fwf_positions(c(1, 30), c(10, 42), c("name", "ssn")))
# 4. Named arguments with start and end positions
read_fwf(fwf_sample, fwf_cols(name = c(1, 10), ssn = c(30, 42)))
# 5. Named arguments with column widths
read_fwf(fwf_sample, fwf_cols(name = 20, state = 10, ssn = 12))
```

### 4. Sometimes strings in a CSV file contain commas. To prevent them from causing problems they need to be surrounded by a quoting character, like " or '. By convention, read_csv() assumes that the quoting character will be ", and if you want to change it you’ll need to use read_delim() instead. What arguments do you need to specify to read the following text into a data frame? ```"x,y\n1,'a,b'"```

In this example, the string is surrounded by the quoting character '. This is not what read_csv() default assumes. In the documentation it looks like you can specify the quote argument for both read_csv() and read_delim(). Below I read in the text using both methods, specifying ```quote = "\'"```.

```{r}
read_csv ("x,y\n1,'a,b'", quote = "\'")
read_delim("x,y\n1,'a,b'", delim = ",", quote = "\'")
```

### 5. Identify what is wrong with each of the following inline CSV files. What happens when you run the code?

I have annotated the code below with the problems for each of the inline CSV files. The output should be displayed if you are viewing the rendered ```.md file``` (you won't see the output if this is a ```.Rmd``` file).

```{r}
# There are not enough column names to go with the amount of columns in the data.
read_csv("a,b\n1,2,3\n4,5,6")
# Mismatched numbers of columns again. The first row only has 2, whereas the 2nd row has 4, and the header only has 3.
read_csv("a,b,c\n1,2\n1,2,3,4")
# Two header columns, and one column of data. Also the "1" is still being read in as an integer.
read_csv("a,b\n\"1")
# Since there are both integer values and character values in the same column, both columns are defined as character.
read_csv("a,b\n1,2\na,b")
# I assume ";" was meant to be the delimiter. The csv only has one observation. Use read_delim("a;b\n1;3", ";") instead.
read_csv("a;b\n1;3")
```

## 11.3 Parsing a vector

Parsing vectors or files can be useful to convert variables to their appropriate classes. For example, if a column of integer values was read in as characters, we can convert the data back into integers using parse_integer(). Below are the provided examples of use:

```{r}
str(parse_logical(c("TRUE", "FALSE", "NA")))
str(parse_integer(c("1", "2", "3")))
str(parse_date(c("2010-01-01", "1979-10-14")))
```

Below are the other examples using parsing functions. Using the problems() function on a parsed vector seems especially useful!

```{r}
# can specify na values if present in data
parse_integer(c("1", "231", ".", "456"), na = ".")

x <- parse_integer(c("123", "345", "abc", "123.45"))
x
problems(x)
```

### Parsing Numbers

Reading in data obtained from outside the US is also tricky since there are different conventions used to display numerical data. For example, using "." instead of "," to mark decimal places or groupings. The functions have arguments that allow you to specify these marks. The examples provided in the book are below. parse_number() also ignores non-numerical symbols such as $ or %. However, parse_double does not seem to have this feature.

```{r}
# you can specify the decimal mark symbol if needed
parse_double("1.23")
parse_double("1,23", locale = locale(decimal_mark = ","))

# parse_number() ignores non-numerical symbols
parse_number("$100")
parse_number("20%")
parse_number("It cost $123.45")

# You can also specify grouping marks if needed.
# Used in America
parse_number("$123,456,789")
# Used in many parts of Europe
parse_number("123.456.789", locale = locale(grouping_mark = "."))
# Used in Switzerland
parse_number("123'456'789", locale = locale(grouping_mark = "'"))
```

### Parsing Strings

You can use parse_character() to parse strings. Each character in a string is encoded, and you can specify the encoding as an argument in parse_character(). To guess the encoding for a particular string you are parsing, you can use guess_encoding().

```{r}
x1 <- "El Ni\xf1o was particularly bad this year"
x2 <- "\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd"

parse_character(x1, locale = locale(encoding = "Latin1"))
parse_character(x2, locale = locale(encoding = "Shift-JIS"))

guess_encoding(charToRaw(x1))
guess_encoding(charToRaw(x2))
```

### Parsing Factors

To parse a vector of factors, you can use parse_factor(), specifying the levels that you are expecting to see. If a value in the vector does not exist in the levels argument, an error is returned.

```{r}
fruit <- c("apple", "banana")
parse_factor(c("apple", "banana", "banana"), levels = fruit)
parse_factor(c("apple", "banana", "bananana"), levels = fruit)
```


### Parsing Dates, Date-times, and Times

There are three types of parsers for these purposes which spit out a combination of date, time, or date-time. Below are the provided examples from the book for each of the parsers.

```{r}
# date-time
# requires input as year, month, day (mandatory), time-(optional)-hour, minute, second, 
parse_datetime("2010-10-01T2010")
parse_datetime("20101010")
```

```{r}
# date - year, month, day
# expects a four digit year, a - or /, the month, a - or /, then the day
parse_date("2010-10-01")
```

```{r}
#  expects the hour, :, minutes, optionally : and seconds, and an optional am/pm specifier:
library(hms)
parse_time("01:10 am")
parse_time("01:10 pm")
parse_time("20:10:01")
```

You can also create your own date-time format. There are many parameters you can specify for your date-time "key". See ?parse_date for the options. Depending on how you set up the "key", you may parse different dates from one set of numbers (book example below).

```{r}
# different dates are parsed depending on the key that you provide
parse_date("01/02/15", "%m/%d/%y")
parse_date("01/02/15", "%d/%m/%y")
parse_date("01/02/15", "%y/%m/%d")

```

Last, as with parsing numbers, different countries may have different date formats. You can solve this by specifying the local argument, as we did with parse_integer().

```{r}
parse_date("1 janvier 2015", "%d %B %Y", locale = locale("fr"))
```

## 11.3.5 Exercises

### 1. What are the most important arguments to locale()?

If you are using locale() for parse_number(), then the most important arguments are ```decimal_mark``` and ```grouping_mark```. For parse_character(), you should specify ```encoding```. For parse_date(), you should specify the region using the appropriate characters.

### 2. What happens if you try and set decimal_mark and grouping_mark to the same character? What happens to the default value of grouping_mark when you set decimal_mark to “,”? What happens to the default value of decimal_mark when you set the grouping_mark to “.”?

When you try to set them to the same character, you get an error: ```Error: `decimal_mark` and `grouping_mark` must be different``` The default grouping_mark becomes '.' if decimal_mark is set to ','.

```{r}
# parse_number("1,234.567", locale = locale(grouping_mark = '.', decimal_mark = '.')) # This Errors!
parse_number("1.234,567", locale = locale(decimal_mark = ','))
parse_number("1.234,567", locale = locale(grouping_mark = '.'))
```

### 3. I didn’t discuss the date_format and time_format options to locale(). What do they do? Construct an example that shows when they might be useful.

The date_format and time_format specify the date and time formats for the parse function, which are by default ```date_format = "%AD"``` and ```time_format = "%AT"```. From the readr vignette, for date_format, "The default value is %AD which uses an automatic date parser that recognises dates of the format Y-m-d or Y/m/d." For time_format, "The default value is %At which uses an automatic time parser that recognises times of the form H:M optionally followed by seconds and am/pm." I could see this useful to specify a custom date_format for american dates, which are often entered as m-d-Y instead of the default Y-m-d. The full four year date is also often truncated to the last 2 digits, which might result in an error without specifying it in date_format. Below is an example.

```{r}
# today's american date parsed incorrectly using default date_format (throws error)
parse_date("05/24/18")
# today's american date parsed correctly by specifying date_format
parse_date("05/24/18", locale = locale(date_format = "%m/%d/%y"))
```

### 4. If you live outside the US, create a new locale object that encapsulates the settings for the types of file you read most commonly.

I live in the US, but for practice purposes lets say I move to Colombia. I might have to commonly read in old files that are not UTF-8 encoded, but latin1 encoded. The decimal marks are also "," instead of ".". Below is an example locale object for these requirements.

```{r}
locale(date_names = "es", decimal_mark = ",", encoding = "latin1")
```

### 5. What’s the difference between read_csv() and read_csv2()?

Base on the documentation, read_csv2() uses semicolons ";" as separators, instead of ",". read_csv2() would ideally be used if the comma "," is used as a decimal point within the file, which would mess up read_csv(). Below is an example:

```{r}
# messed up because "," is used as a decimal point
read_csv("a;b\n1,0;2,0")
# using read_csv2 fixes the problem
read_csv2("a;b\n1,0;2,0")
```

### 6. What are the most common encodings used in Europe? What are the most common encodings used in Asia? Do some googling to find out.

I got the info from the ?stringi::stri_enc_detect documentation.

Common encodings used in Europe are: ISO-8859-1, ISO-8859-2, windows-1252, ISO-8859-7
Common encodings used in Asia are: Shift_JIS, ISO-2022-JP, ISO-2022-CN, ISO-2022-KR, GB18030, EUC-JP, EUC-KR

UTF-8 is widely popular now, and you can also use guess_encoding() if you are unsure what encoding to use. There is also a lot of info about encoding on Wikipedia.

### 7. Generate the correct format string to parse each of the following dates and times:

My answers are in the R code below. Helpful descriptions for the format string paramters are found at ?parse_datetime.

```{r}
d1 <- "January 1, 2010"
parse_date (d1, "%B%e%*%Y")

d2 <- "2015-Mar-07"
parse_date(d2, "%Y-%b-%d")

d3 <- "06-Jun-2017"
parse_date(d3, "%d-%b-%Y")

d4 <- c("August 19 (2015)", "July 1 (2015)")
parse_date(d4, "%B %d (%Y)")

d5 <- "12/30/14" # Dec 30, 2014
parse_date(d5, "%m/%d/%y")

t1 <- "1705"
parse_time(t1, "%H%M")

t2 <- "11:15:10.12 PM"
parse_time(t2, "%I:%M:%OS %p")
```

## 11.4 Parsing a file

The parsers that we learned about in the previous section are automatically applied by readr when reading in a file using read_csv() or other reading functions. These guess the type of data in each column being read in, using a combination of guess_parser() and parse_guess() on the first 1000 rows of observations.
```{r}
guess_parser(c("TRUE", "FALSE"))
parse_guess("2010-10-10")
```

There are usually a lot of issues when parsing a large, unorganized file. readr has a "challenge" example that displays some of the issues that arise:

```{r}
challenge <- read_csv(readr_example("challenge.csv"))
```

Since the default only looks at the first 1000 rows, we can run into issues if the first 1000 rows or more have troubling characteristics. Here there are many issues displayed in the output after attempting to read the file in. It is very helpful that the error output displays what the function attempted to do. We see that it attempted to parse column x using col_integer(), and column y using col_character(). We can see more details by using problems():

```{r}
problems(challenge)
tail(challenge)
typeof(challenge$x)
typeof(challenge$y)
unique(challenge$y[1:1000])
```

The values for column x after row 1000 seem to be doubles, rather than integers. We can fix this by changing the default parsing function from col_integer() to col_double(). We also observe that column y contains date values, but the default type was character, since the first 1000 values were NA. We can fix this by changing col_character() to col_date().

```{r}
challenge <- read_csv(
  readr_example("challenge.csv"), 
  col_types = cols(
    x = col_double(),
    y = col_date()
  )
)
tail(challenge)
typeof(challenge$x)
typeof(challenge$y)
```

Hadley recommends that we should always examine the output of the read_() function and re-specify the col_parsers to match what is appropriate for the data. One strategy around this that he describes, which I think would probably be more straightforward if you have many, many columns of data, is to read everything in as a character, then use type_convert() on the table to convert to the appropriate types. We can see in the example below that type_convert() properly converts column x to double and column y to date formats.

```{r}
challenge2 <- read_csv(readr_example("challenge.csv"), 
  col_types = cols(.default = col_character())
)
challenge2
type_convert(challenge2)
```

The functions read_lines() and read_file() also seem useful to read in the raw lines or unstructured text of a file, in order to better understand the type of data contained. Unless you want to manipulate the strings or extract data using regexes, it might be more efficient to use ```less``` in your terminal to view the data file, rather than read_file().

```{r}
# use read_lines to read individual lines of the file
head(read_lines(readr_example("challenge.csv")))
#use read_file to read the entire file in as one string
substr(read_file(readr_example("challenge.csv")), 1, 100)
```

## 11.5 Writing to a file

The functions write_csv() and write_tsv() are useful functions to write data.frames or tibbles in R to files. When writing files, it is important to use UTF-8 encoding for strings and save dates/date-times in ISO8601 format. There is also a special function for writing to excel: write_excel_csv(). write_rds() will save the actual R object containing the data frame, so that if you load the ```.rds``` using read_rds() you can access the data as it was at the time of save. Think of this as using save() in base R to save a ```.Robj``` for a variable you want to keep track of. You could also use write_feather() from the ```feather``` package to save the data in a format accessible by other programming languages, or read it back into R using read_feather().

```{r}
# write_csv(challenge, "challenge.csv")
# write_rds(challenge, "challenge.rds")
# library(feather)
# write_feather(challenge, "challenge.feather")
```

When executing the above write_csv() or write_rds(), the files will appear in your working directory, which if you are using an R notebook, conveniently is where your ```.Rmd``` file is being kept!

There are also other types of files that might be read in. You can use haven for reading in SPSS, Stata, and SAS files, readxl for excel files, DBI to run SQL queries against databases (returns data frame), jsonlite for json, and xml2 for XML.

# Chapter 12 Tidy Data

Below is the example provided by the book of the same data presented in 4 different ways. One way is tidy, the others are not!

```{r}
table1
table2
table3
# data spread across 2 tables
table4a # cases
table4b # population
```

The three rules that must be satisfied for data to be "tidy" are: 

* Each variable must have its own column.
* Each observation must have its own row.
* Each value must have its own cell.

Using these rules, the a tidy dataset from above is ```table1```. The packages in the tidyverse such as ggplot2, dplyr, etc. are designed to work with tidy data, so we must learn how to reorganize data and clean it up in the event that we receive a dataset that is not tidy (which is very often).

## 12.2.1 Exercises

### 1. Using prose, describe how the variables and observations are organised in each of the sample tables.

* In table1, each of the variables are organized into columns, and each observation is in a separate row. Each row presents a unique combination of variables and respective observations, and each value has its own cell. This is considered a tidy dataset.
* In table2, each observation is in a separate row, but each variable does not have its own column. This is not a tidy dataset. To "tidy" this data, population and cases should be separated into individual rows, based on the count column. Count is not considered a variable. Rather, it is considered a value.
* In table 3, population and cases have been combined into a single column, "rate". Note that this is **not** the actual value of the cases divided by the population (as you might obtain using a mutate()), rather it is a character listing the two separate items! If we had not known there were two variables combined into one column, we might have called this a tidy dataset. I would find a way to split the rate into cases and population, then use mutate() to calculate the rate.
* In table 4a and 4b, the data for population and cases have been split into two separate tables. The date variable is spread across the columns, and each row of each table represents an individual observation. The two tables have the same country column, and look like something you would see when working with relational data.

### 2. Compute the rate for table2, and table4a + table4b. You will need to perform four operations:

* Extract the number of TB cases per country per year.
* Extract the matching population per country per year.
* Divide cases by population, and multiply by 10000.
* Store back in the appropriate place.

Which representation is easiest to work with? Which is hardest? Why?

I will try doing the above operations without using tidyr functions. 

For table2, we can first filter for the TB cases using filter(), then filter for the population per country using filter() as well. Then we can divide the two, multiply by 10000, and bind the results back to the original table2 by using cbind(). Since there are two rows in table2 corresponding to the same rate value, we must duplicate each rate twice using ```rep( each = 2 )``` before binding to table2. This preserves the original structure of table2, and has the rate stored in the appropriate places. Using the spread() function in tidyr to generate tidy data and then using mutate() is much easier!

For table 4a and 4b, we can extract the 1999 values from each table and perform the rate calculation, and do the same thing for the 2000 values.Then, we can bind these rates back to both table4a and 4b using cbind. This preserves the original structure for both tables.

```{r}
# perform the operations above for table 2a, without using spread()
rate_table2 <- ((filter (table2, type == 'cases')$count / # select cases
  filter (table2, type == 'population')$count) * 10000) %>% # divide by populatoin
  rep(each = 2) # repeat each element twice, maintaining the order of the vector
cbind(table2,rate_table2)

# using spread() and dplyr is more straightforward and results in tidy data.
table2 %>%
  spread(key = type, value = count) %>%
  mutate(rate = (cases / population) * 10000)

# perform the operations above for table 4a and 4b
rate_1999 <- (table4a$`1999` / table4b$`1999`)*10000
rate_2000 <- (table4a$`2000` / table4b$`2000`)*10000
cbind (table4a, rate_1999, rate_2000)
cbind (table4b, rate_1999, rate_2000)


```

### 3. Recreate the plot showing change in cases over time using table2 instead of table1. What do you need to do first?

First, we need to tidy table2 so that we can use it with ggplot2, using the spread() function. Then, we can use the ggplot() function in the same manner as in the book example.

```{r}
table2 %>%
  spread(key = type, value = count) %>%
  ggplot(aes(year, cases)) + 
    geom_line(aes(group = country), colour = "grey50") + 
    geom_point(aes(colour = country))
```

## 12.3.1 Spreading and Gathering

If some column names are not names of variables, but rather values of variables, it will be helful to "gather" these columns into a single column and map the values accordingly. For example, table4a and table4b has values for 1999 and 2000 split across multiple columns. To gather the values, use gather() by specifying the names of the columns to gather, and the names of the variable that is being gathered (key) and the value (value) that is being redistributed.

```{r}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")

table4b %>% 
  gather(`1999`, `2000`, key = "year", value = "population")
```

When an observation is spread across multiple rows, as in table2, you can spread the data into multiple columns such that each row contains data from a single observation. In table2, each observation is considered a country in a year. Cases and populations for each country/year observation can be split from the type column into their own columns.

```{r}
spread(table2, key = type, value = count)
```

## 12.3.3 Exercises

### 1. Why are gather() and spread() not perfectly symmetrical? Carefully consider the following example:
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks
stocks %>% 
  spread(year, return)
stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`)
```
(Hint: look at the variable types and think about column names.)

There are a couple differences between the unmodified stocks tibble and the spread and gathered tibble. The spread and gathered stocks tibble has the columns ordered differently. Also, the the types of each column have changed. The original tibble had 3 columns of type <dbl> whereas the spread and gathered tibble has changed the year column to <chr>. The column names can also be changed by spreading and gathering, since they have to be specified as arguments in gather().

Both spread() and gather() have a convert argument. What does it do?

If set to TRUE (the default is FALSE), then the function type.convert() with asis = TRUE will be run on each of the new columns in the spread() or gather() output. This will try to convert the columns to variables of the appropriate type. For example, using convert = TRUE on the example above causes the year column to become an <int> rather than <chr>.

```{r}
stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`, convert = TRUE)
```

### 2. Why does this code fail?
```{r}
# table4a %>% 
#   gather(1999, 2000, key = "year", value = "cases")
#> Error in combine_vars(vars, ind_list): Position must be between 0 and n
```

The code fails because the variables passed into gather(), 1999 and 2000, are not written correctly. They should be encased in backticks, like this: `1999` and `2000`. The code below works.

```{r}
table4a %>% 
   gather(`1999`, `2000`, key = "year", value = "cases")
```

### 3. Why does spreading this tibble fail? How could you add a new column to fix the problem?
```{r}
people <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
# spreading fails
#spread(people, key, value)

people2 <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods2",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
spread(people2, key, value)
people3 <- tribble(
  ~name,             ~key,    ~value, ~index,
  #-----------------|--------|------|------
  "Phillip Woods",   "age",       45, 1,
  "Phillip Woods",   "height",   186, 1,
  "Phillip Woods",   "age",       50, 2,
  "Jessica Cordero", "age",       37, 3,
  "Jessica Cordero", "height",   156, 3
)
# works now
spread(people3, key, value)
```

Spreading the provided tribble fails because there is a duplicate value for the variables name and key. If I modify the second "Phillip Woods" value to "Phillip Woods2", the spread() function works. If I add another column named "index" that identifies the unique sets of age and height, spread() also works. In any case, spread() requires a unique combination of variables to spread against.

### 4. Tidy the simple tibble below. Do you need to spread or gather it? What are the variables?
```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
```

I would need to gather the data. We can combine the male and female designation into a column called "gender", and gather the 4 numerical observations into a column called "cases". Below is the code required to do this.

```{r}
preg %>%
  gather(male, female, key = "gender", value = "cases")
```


## 12.4 Separating and uniting

In table3, the data for population and cases are combined into one column, rate. To separate them, we can use the separate() function in tidyr. In the opposite case, the function unite() can be used to combine an observation split across multiple columns. Do not confuse unite() with gather()!

Separate() separates one column into multiple columns, by splitting based on a separator character. If the separator is the only non-numerical character for each observation, it does not have to be specified.

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
```

Similar to spread() and gather(), separate() has a convert argument that can try to convert the columns to the appropriate types. You can also specify the index to separate the values on, using a number instead of string value for the sep argument.

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
```

Unite() is the opposite of separating, and will combine two columns together using a default underscore "_". You can change the underscore to your choice of separator, using and empty "" if you do not wish to separate the values. The convert argument does not seem to work with this function in the example below.

```{r}
# use default separator for unite()
table5 %>%
  unite(new, century,year)
# specify sep = "" to combine the two columns with no separation
table5 %>% 
  unite(new, century, year, sep = "")
```

## 12.4.3 Exercises
### 1. What do the extra and fill arguments do in separate()? Experiment with the various options for the following two toy datasets.
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"))

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"))
```

In the examples above, the tibbles created are 1 column, 3 rows, with strings as observations. Within each string, commas separate the letters. The separate(x, c("one", "two", "three")) will try to split each string using the comma into three columns. However, in the first tibble, there is a row with four values separated by commas (an extra value). This is where the ```extra``` argument comes into play. 

The documentation for extra states that:
* "warn" (the default): emit a warning and drop extra values.
* "drop": drop any extra values without a warning.
* "merge": only splits at most length(into) times

For the second tibble, there are only two values separated by a comma in the second row (a missing value). This is where the ```fill``` argument comes into play.

The documentation for fill states that:

* "warn" (the default): emit a warning and fill from the right
* "right": fill with missing values on the right
* "left": fill with missing values on the left

The default for both extra and fill is "warn", which just provides a warning about the missing or extra value. Below I will try the other two options on the appropriate tibles.

```{r}
# tibble with extra value
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "drop")
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "merge")

# tible with missing value
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "right")
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "left")
```

### 2. Both unite() and separate() have a remove argument. What does it do? Why would you set it to FALSE?

The documentation states that "If TRUE, remove input column from output data frame." The column with the values that were split is by default removed from the resulting table, but if this parameter is set to FALSE, the column remains. You might want to set it to FALSE if you want to have a record of what the original column looked like or want to separate the column a different way. Below is an exmaple of setting remove = FALSE on table3. We can see that the rate column still exists in the output.

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE, remove = FALSE)
```

### 3. Compare and contrast separate() and extract(). Why are there three variations of separation (by position, by separator, and with groups), but only one unite?

extract() is a version of separate() except you can specify which groups to include in the final output using regexes. For example, if you run the default extract() on a data frame, it will first separate based on a non-numeric delimiter, then only place the number of groups you specify in the into argument, starting from the left. separate() will throw an error if you specify less than the number of groups for the ```into``` argument, and by default want to return all the groups created.

There are three variations of separation because the output can be flexible based on the type of separation used (there are fewer constraints on how a value can be split up). There is only one unite because combining values together has more constraints. Unless you can take bits and pieces of different variables to unite (and define how that occurs), the only thing you can define is the delimiter that is used when uniting the data.

```{r}
# separate errors if you try to separate into less columns than groups, but can still function.
table3 %>%
  separate(rate, into = c("cases"))
# extract does not error
table3 %>%
  extract(rate, into = c("cases"))
```

## 12.5 Missing Values

Values in your dataset can be missing explicitly, in which they are specified as NA, or implicitly, in which they are not specified at all. The example provided is:

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
stocks
```

In which there is an explicitly missing value for year 2015 quarter 4, and an implicitly missing value for year 2016 quarter 1 (which is just not present in the data).

We can expose the implicitly missing value by spreading the data:

```{r}
stocks %>% 
  spread(year, return)
```

Now we see two explicitly missing values. We can also omit NA values entirely by using na.omit() or by specifying na.rm = T in a spread and gather combination.

```{r}
# get rid of NA values using na.omit()
na.omit(stocks)
# get rid of NA values using spread and gather( na.rm = T )
stocks %>% 
  spread(year, return) %>% 
  gather(year, return, `2015`:`2016`, na.rm = TRUE)
```

A very useful tool is the complete() function in tidyr! It will look for all combinations of the variables specified and find implicitly missing values in the data, and turn them explicit! This is much cleaner-looking than spreading and gathering.

```{r}
stocks %>% 
  complete(year, qtr)
```


```{r}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
treatment
treatment %>% 
  fill(person)
```



##12.5.1 Exercises

### 1. Compare and contrast the fill arguments to spread() and complete().

### 2. What does the direction argument to fill() do?




## 12.6.1 Exercises

### 1. In this case study I set na.rm = TRUE just to make it easier to check that we had the correct values. Is this reasonable? Think about how missing values are represented in this dataset. Are there implicit missing values? What’s the difference between an NA and zero?

### 2. What happens if you neglect the mutate() step? (mutate(key = stringr::str_replace(key, "newrel", "new_rel")))

### 3. I claimed that iso2 and iso3 were redundant with country. Confirm this claim.

### 4. For each country, year, and sex compute the total number of cases of TB. Make an informative visualisation of the data.






# Chapter 13 Relational Data

## 13.2.1 Exercises

### 1. Imagine you wanted to draw (approximately) the route each plane flies from its origin to its destination. What variables would you need? What tables would you need to combine?

### 2. I forgot to draw the relationship between weather and airports. What is the relationship and how should it appear in the diagram?

### 3. weather only contains information for the origin (NYC) airports. If it contained weather records for all airports in the USA, what additional relation would it define with flights?

### 4. We know that some days of the year are “special”, and fewer people than usual fly on them. How might you represent that data as a data frame? What would be the primary keys of that table? How would it connect to the existing tables?


## 13.3.1 Exercises

### 1. Add a surrogate key to flights.

### 2. Identify the keys in the following datasets

* Lahman::Batting,
* babynames::babynames
* nasaweather::atmos
* fueleconomy::vehicles
* ggplot2::diamonds
(You might need to install some packages and read some documentation.)

### 3. Draw a diagram illustrating the connections between the Batting, Master, and Salaries tables in the Lahman package. Draw another diagram that shows the relationship between Master, Managers, AwardsManagers. How would you characterise the relationship between the Batting, Pitching, and Fielding tables?



## 13.4.6 Exercises

### 1. Compute the average delay by destination, then join on the airports data frame so you can show the spatial distribution of delays. Here’s an easy way to draw a map of the United States:
```{r}
library(nycflights13)
airports %>%
  semi_join(flights, c("faa" = "dest")) %>%
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

(Don’t worry if you don’t understand what semi_join() does — you’ll learn about it next.)
You might want to use the size or colour of the points to display the average delay for each airport.

### 2. Add the location of the origin and destination (i.e. the lat and lon) to flights.

### 3. Is there a relationship between the age of a plane and its delays?

### 4. What weather conditions make it more likely to see a delay?

### 5. What happened on June 13 2013? Display the spatial pattern of delays, and then use Google to cross-reference with the weather.


## 13.5.1 Exercises
### 1. What does it mean for a flight to have a missing tailnum? What do the tail numbers that don’t have a matching record in planes have in common? (Hint: one variable explains ~90% of the problems.)

### 2. Filter flights to only show flights with planes that have flown at least 100 flights.

### 3. Combine fueleconomy::vehicles and fueleconomy::common to find only the records for the most common models.

### 4. Find the 48 hours (over the course of the whole year) that have the worst delays. Cross-reference it with the weather data. Can you see any patterns?

### 5. What does anti_join(flights, airports, by = c("dest" = "faa")) tell you? What does anti_join(airports, flights, by = c("faa" = "dest")) tell you?

### 6. You might expect that there’s an implicit relationship between plane and airline, because each plane is flown by a single airline. Confirm or reject this hypothesis using the tools you’ve learned above.


