---
title: "R for Data Science Walkthrough Chapters 4-8"
author: "Erick Lu"
output: 
  github_document:
    html_preview: false
    toc: true
    toc_depth: 2
    fig_width: 3.5
    fig_height: 2.5
    dev: jpeg
---

This my walkthrough for the book: _R for Data Science_ by Hadley Wickham and Garrett Grolemund. It contains my answers to their exercises and some of my own notes and data explorations. Here I will go through chapters 4-8.

# Chapter 4

# Chapter 5

The data that we will work with in chapter 5 is from the nycflights13 package.

```{r}
library(nycflights13)
library(tidyverse)

flights
```

## 5.2 Filter

filter() will subset obervations based on their values. I think it works a lot like the which() function in base R (ie: data[which(data$variable > value),] ). Below is a way to do the same thing using either filter() or base R which().

```{r}
# tidyverse filter() output
filter(flights, month == 1, day == 1)

# base R way to get the same output
flights[which(flights$month ==1 & flights$day ==1),]
```









